---
title: "Guided tutorial to Differential Expression Analysis using COTAN"
author: 
- name: "Silvia Giulia Galfrè"
  affiliation: "Department of Computer Science, University of Pisa"
- name: "Marco Fantozzi"
  affiliation: "University of Parma"

package: COTAN
output:
  BiocStyle::html_document:
    keep_md: true
vignette: >
  %\VignetteIndexEntry{DEA using COTAN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Preamble

```{r knitr options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7L,
  fig.height = 5L,
  dpi = 100
)

if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
} else if (requireNamespace("Cairo", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "cairo_png")
} else {
  knitr::opts_chunk$set(dev = "png")
}
```


```{r preamble, message=FALSE, warning=FALSE}
library(COTAN)
library(zeallot)

# necessary to solve precedence of overloads
conflicted::conflict_prefer("%<-%", "zeallot")

# enable multi-processing (not on Windows)
prevOptState <- options(parallelly.fork.enable = TRUE)
```

# Introduction

This tutorial shows the available tools in COTAN to help with genes' clustering

### Setup

Define a directory where the data and the output will be stored

```{r setup-and-log}
dataDir <- tempdir()
print(dataDir)

outDir <- dataDir

# Log-level 2 was chosen to showcase better how the package works
# In normal usage a level of 0 or 1 is more appropriate
setLoggingLevel(2L)

# This file will contain all the logs produced by the package
# as if at the highest logging level
setLoggingFile(file.path(outDir, "vignette_DEA.log"))
```


## Retrieving the data-set

Download the data-set for `r "mouse cortex E17.5"` from
`r "Yuzwa et al."` (2017)

```{r download-data}
GEO <- "GSM2861514"

dir.create(file.path(dataDir, GEO), showWarnings = FALSE)

datasetFileName <-
  file.path(dataDir, GEO, "GSM2861514_E175_All_Cells_DGE.txt.gz")

# retries up to 5 times to get the dataset
attempts <- 0L
repeat {
  if (!file.exists(datasetFileName)) {
    GEOquery::getGEOSuppFiles(
      GEO,
      makeDirectory = TRUE,
      baseDir = dataDir,
      filter_regex = base::basename(datasetFileName),
      fetch_files = TRUE
    )
  }

  attempts <- attempts + 1L
  if (attempts >= 5L) {
    break
  }
}

assertthat::assert_that(file.exists(datasetFileName))

rawDataset <- read.csv(datasetFileName, sep = "\t", row.names = 1L)

print(dim(rawDataset))
```

### Create the COTAN object

Initialize the `COTAN` object with the row count table and
the metadata from the experiment.

```{r create-cotan-object}
cond <- "mouse_cortex_E17.5"

obj <- COTAN(raw = rawDataset)
obj <-
  initializeMetaDataset(
    obj,
    GEO = GEO,
    sequencingMethod = "Drop_seq",
    sampleCondition = cond
  )

logThis(paste0("Condition ", getMetadataElement(obj, datasetTags()[["cond"]])),
        logLevel = 1L)
```

Assign to each cell its origin

```{r add-origin-to-the-cotan-object}
# mark cells origin
data(vignette.cells.origin)
head(vignette.cells.origin, 18)

obj <- 
  addCondition(obj, condName = "origin", conditions = vignette.cells.origin)
```


### Drop low quality cells

```{r drop-low-quality-cells}
# use previously established results to determine
# which cells were dropped in the cleaning stage
data(vignette.split.clusters)

cellsToDrop <-
  getCells(obj)[!(getCells(obj) %in% names(vignette.split.clusters))]

obj <- dropGenesCells(obj, cells = cellsToDrop)

# Log the remaining number of cells

logThis(paste("n cells", getNumCells(obj)), logLevel = 1L)

table(getCondition(obj, condName = "origin"))

rm(vignette.split.clusters)
```

# Establish Uniform Transcript cells' clusters

## COTAN analysis

In this part, the `COTAN` model is calibrated:
all the contingency tables are computed and used to get
the statistics necessary to `COEX` evaluation and storing

```{r calculate-and-store-coex-matrix}
obj <-
  proceedToCoex(
    obj,
    calcCoex = FALSE,
    optimizeForSpeed = TRUE,
    cores = 3L,
    deviceStr = "cuda"
  )
```


## Uniform Clustering

It is possible to obtain a cells' *clusterization* based on the concept of 
*uniformity* of expression of the genes across the cells in each cluster.
That is the clusters satisfy the null hypothesis of the `COTAN` model:
the genes expression is not dependent on the cell in consideration.

### First UT clusterization

Loading pre-calculated *clusterization* from data

```{r load-pre-calculated-clusterization_split}
data("vignette.split.clusters", package = "COTAN")
splitClusters <- vignette.split.clusters

# explicitly calculate the DEA of the clusterization to store it
splitCoexDF <- DEAOnClusters(obj, clusters = splitClusters)

obj <-
  addClusterization(
    obj,
    clName = "split",
    clusters = splitClusters,
    coexDF = splitCoexDF,
    override = FALSE
  )
```


### Clusterization statistics

It is possible to extract some statistics about a stored *clusterization*:

```{r extract-clusterization-summary_split}
c(summaryData, summaryPlot) %<-%
  clustersSummaryPlot(
    obj,
    clName = "split",
    plotTitle = "split summary",
    condName = "origin"
  )

head(summaryData, n = 10L)

plot(summaryPlot)
```
The `ExpGenes` column contains the number of genes that are expressed in any
cell of the relevant *cluster*, while the `ExpGenes25` column contains the
number of genes expressed in at the least 25% of the cells of the relevant
*cluster*

### Relevance of marker genes for the `split` clusters

It is possible to visualize how relevant are some *marker genes* for the
*clusters* comprising a given *clusterization*

```{r define-layers-specific-genes}
# these are some genes associated to each cortical layer
layersGenes <- list(
  "L1"   = c("Reln",   "Lhx5"),
  "L2/3" = c("Cux1",   "Satb2"),
  "L4"   = c("Rorb",   "Sox5"),
  "L5/6" = c("Bcl11b", "Fezf2"),
  "Prog" = c("Hes1",    "Vim")
)
```


```{r clusterization-data-visualization_split}
c(splitHeatmap, splitScoreDF, splitPValueDF) %<-%
  clustersMarkersHeatmapPlot(
    obj,
    groupMarkers = layersGenes,
    clName = "split",
    kCuts = 2L,
    adjustmentMethod = "holm"
  )

ComplexHeatmap::draw(splitHeatmap)
```

In the above graph, it is possible to see that the found clusters align
well to the expression of the layers' genes.

```{r eval=FALSE}
enrichment <- geneSetEnrichment(clustersCoex = coexDF,
                                groupMarkers = groupMarkers)

```


### Merged UT clusterization

Loading pre-calculated *clusterization* from data

```{r load-pre-calculated-clusterization_merge}
data("vignette.merge.clusters", package = "COTAN")
mergedClusters <- vignette.merge.clusters[getCells(obj)]

# explicitly calculate the DEA of the clusterization to store it
mergedCoexDF <- DEAOnClusters(obj, clusters = mergedClusters)

obj <-
  addClusterization(
    obj,
    clName = "merge",
    clusters = mergedClusters,
    coexDF = mergedCoexDF,
    override = FALSE
  )

# this gives immediate visual information about which clusters were merged
table(splitClusters, mergedClusters)
```

```{r extract-clusterization-summary_merge}
c(summaryData, summaryPlot) %<-%
  clustersSummaryPlot(
    obj,
    clName = "split",
    plotTitle = "split summary",
    condName = "origin"
  )

head(summaryData, n = 10L)

plot(summaryPlot)
```

### Relevance of marker genes for the `merge` clusters

Again, it is possible to visualize how relevant are the *marker genes* from
above:

```{r  clusterization-data-visualization_merge}
c(mergeHeatmap, mergeScoresDF, mergePValuesDF) %<-%
  clustersMarkersHeatmapPlot(
    obj,
    groupMarkers = layersGenes,
    clName = "merge",
    condNameList = list("origin")
  )

ComplexHeatmap::draw(mergeHeatmap)
```


## Vignette clean-up stage

The next few lines are just to clean.

```{r clean-up}
if (file.exists(file.path(outDir, paste0(cond, ".cotan.RDS")))) {
  # delete file if it exists
  file.remove(file.path(outDir, paste0(cond, ".cotan.RDS")))
}
if (file.exists(file.path(outDir, paste0(cond, "_times.csv")))) {
  # delete file if it exists
  file.remove(file.path(outDir, paste0(cond, "_times.csv")))
}
if (dir.exists(file.path(outDir, cond))) {
  unlink(file.path(outDir, cond), recursive = TRUE)
}
# if (dir.exists(file.path(outDir, GEO))) {
#   unlink(file.path(outDir, GEO), recursive = TRUE)
# }

# stop logging to file
setLoggingFile("")
file.remove(file.path(outDir, "vignette_uniform_clustering.log"))

options(prevOptState)
```

```{r closure}
Sys.time()

sessionInfo()
```
