---
title: "Guided tutorial to Differential Expression Analysis using COTAN"
author: 
- name: "Silvia Giulia Galfrè"
  affiliation: "Department of Computer Science, University of Pisa"
- name: "Marco Fantozzi"
  affiliation: "University of Parma"

package: COTAN
output:
  BiocStyle::html_document:
    keep_md: true
vignette: >
  %\VignetteIndexEntry{DEA using COTAN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Preamble

```{r knitr options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7L,
  fig.height = 5L,
  dpi = 100,
  fig.crop = FALSE
)

if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
} else if (requireNamespace("Cairo", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "cairo_png")
} else {
  knitr::opts_chunk$set(dev = "png")
}
```


```{r preamble, message=FALSE, warning=FALSE}
library(COTAN)
library(zeallot)

# necessary to solve precedence of overloads
conflicted::conflict_prefer("%<-%", "zeallot")

# enable multi-processing (not on Windows)
prevOptState <- options(parallelly.fork.enable = TRUE)
```

# Introduction

This tutorial shows the available tools in COTAN to help with genes' clustering

### Setup

Define a directory where the data and the output will be stored

```{r setup-and-log}
dataDir <- file.path(tempdir(), "COTAN_vignette_data")
dir.create(dataDir, recursive = TRUE, showWarnings = FALSE)

print(dataDir)

outDir <- dataDir

# Log-level 2 was chosen to showcase better how the package works
# In normal usage a level of 0 or 1 is more appropriate
setLoggingLevel(2L)

# This file will contain all the logs produced by the package
# as if at the highest logging level
setLoggingFile(file.path(outDir, "vignette_DEA.log"))
```


## Retrieving the data-set

Download the data-set for `r "mouse cortex E17.5"` from
`r "Yuzwa et al."` (2017)

```{r download-data}
GEO <- "GSM2861514"

dir.create(file.path(dataDir, GEO), showWarnings = FALSE)

datasetFileName <-
  file.path(dataDir, GEO, "GSM2861514_E175_All_Cells_DGE.txt.gz")


# retries up to 5 times to get the dataset
attempts <- 0L
maxAttempts <- 5L
ok <- FALSE

while (attempts < maxAttempts && !ok) {
  attempts <- attempts + 1L

  if (!file.exists(datasetFileName)) {
    res <- try(
      GEOquery::getGEOSuppFiles(
        GEO = GEO,
        makeDirectory = TRUE,
        baseDir = dataDir,
        filter_regex = base::basename(datasetFileName),
        fetch_files = TRUE
      ),
      silent = TRUE
    )
  }

  ok <- file.exists(datasetFileName)

  if (!ok && attempts < maxAttempts) {
    Sys.sleep(1)
  }
}

assertthat::assert_that(
  ok,
  msg = paste0(
    "Failed to retrieve file '", datasetFileName,
    "' after ", maxAttempts, " attempts."
  )
)

rawDataset <- read.csv(datasetFileName, sep = "\t", row.names = 1L)

print(dim(rawDataset))
```

### Create the COTAN object

Initialize the `COTAN` object with the row count table and
the metadata from the experiment.

```{r create-cotan-object}
cond <- "mouse_cortex_E17.5"

obj <- COTAN(raw = rawDataset)
obj <-
  initializeMetaDataset(
    obj,
    GEO = GEO,
    sequencingMethod = "Drop_seq",
    sampleCondition = cond
  )

logThis(paste0("Condition ", getMetadataElement(obj, datasetTags()[["cond"]])),
        logLevel = 1L)
```

Assign to each cell its origin

```{r add-origin-to-the-cotan-object}
# mark cells origin
data(vignette.cells.origin)
head(vignette.cells.origin, 18)

obj <- 
  addCondition(obj, condName = "origin", conditions = vignette.cells.origin)

rm(vignette.cells.origin)
```


### Drop low quality cells

```{r drop-low-quality-cells}
# use previously established results to determine
# which cells were dropped in the cleaning stage
data(vignette.split.clusters)

cellsToDrop <-
  getCells(obj)[!(getCells(obj) %in% names(vignette.split.clusters))]

obj <- dropGenesCells(obj, cells = cellsToDrop)

# Log the remaining number of cells

logThis(paste("n cells", getNumCells(obj)), logLevel = 1L)

table(getCondition(obj, condName = "origin"))

rm(vignette.split.clusters)
```

## COTAN analysis

In this part, the `COTAN` model is calibrated, but the `COEX` matrix is not
evaluated as it is not strictly necessary

```{r calculate-and-store-coex-matrix}
obj <-
  proceedToCoex(
    obj,
    calcCoex = FALSE,
    optimizeForSpeed = TRUE,
    cores = 3L,
    deviceStr = "cuda"
  )
```


# Differential Expression Analysis

TODO:

### Existing clusterizations

Loading pre-calculated *clusterizations* from data and call the
`DEAOnClusters()`. This method estimates assess the differential expression of genes when partitioning cells into two subsets: the cluster and all other cells in the dataset. For each cluster `COTAN` calculates a *correlation coefficient*
based on contingency tables that counts whether each gene occurs or not in
cells inside or outside of the cluster.

These numbers are organized as one column for each cluster with a row for each
gene and are a measure of the enrichment or depletion of the genes inside the
clusters.

```{r load-pre-calculated-clusterizations}
data("vignette.split.clusters", package = "COTAN")
data("vignette.merge.clusters", package = "COTAN")

# explicitly calculate the DEA of the clusterization to store it
vignette.split.coexDF <- DEAOnClusters(obj, clusters = vignette.split.clusters)
vignette.merge.coexDF <- DEAOnClusters(obj, clusters = vignette.merge.clusters)

obj <-
  addClusterization(
    obj,
    clName = "split",
    clusters = vignette.split.clusters,
    coexDF = vignette.split.coexDF,
    override = FALSE
  )

obj <-
  addClusterization(
    obj,
    clName = "merge",
    clusters = vignette.merge.clusters,
    coexDF = vignette.merge.coexDF,
    override = FALSE
  )

# these will be recovered from the COTAN obj when required
rm(vignette.split.clusters, vignette.split.coexDF)
rm(vignette.merge.clusters, vignette.merge.coexDF)
```

## Manipulate a clusterization

It is possible to improve on the labels in a *clusterization*, so that they
look nicer in plots and lists, for example to ensure all labels have the
**same** length.

Another improvement is to reorder the labels of the *clusterization* so that *near* clusters have *near* labels

```{r}
# COTAn always stores clusterizations as factors.
splitClusters <- getClusters(obj, clName = "split")

# Use the utility factorToVector() to properly decay any named factor
# to a named char array
# splitClusters <- factorToVector(splitClusters)

# In this case the following calls are no-op since the clusterization
# were created by COTAN and so they already made nice and reordered

splitClusters <- niceFactorLevels(splitClusters)

c(splitClusters, splitCoexDF, perm) %<-% 
  reorderClusterization(
    obj,
    clName = "split",
    reverse = FALSE, 
    coexDF = NULL,
    useDEA = TRUE, # will use DEA columns with cosine distance
    distance = NULL,
    hclustMethod = "ward.D2"
  )

mergeClusters <- getClusters(obj, clName = "merge")
```

`COTAN` provides also easy facility to convert a *clusterization* from a
*factor* to a *list* of *array* of cells and vice-versa.
This can come useful if one wants to iterate separately on the cells of each
cluster.

```{r}
splitClustersAsList <- toClustersList(splitClusters)

assertthat::assert_that(length(splitClustersAsList) == nlevels(splitClusters))

for (clName in names(splitClustersAsList)) {
  cluster <- splitClustersAsList[[clName]]
  clRawData <- getRawData(obj)[, cluster, drop = FALSE]
  clRawData <- clRawData[clRawData > 0.0]
  print(paste("Cluster", clName,
              "has an average non-zero expression of", mean(clRawData)))
  rm(clRawData)
}
```

## Relevance of marker genes

It is possible to visualize how relevant are some *marker genes* for the
*clusters* comprising a given *clusterization*

```{r define-layers-specific-genes}
# these are some genes associated to each cortical layer
layersGenes <- list(
  "L1"   = c("Reln",   "Lhx5"),
  "L2/3" = c("Cux1",   "Satb2"),
  "L4"   = c("Rorb",   "Sox5"),
  "L5/6" = c("Bcl11b", "Fezf2"),
  "Prog" = c("Hes1",    "Vim")
)
```



```{r clusterization-data-visualization_split}
c(splitHeatmap, splitScoreDF, splitPValueDF) %<-%
  clustersMarkersHeatmapPlot(
    obj,
    groupMarkers = layersGenes,
    clName = "split",
    kCuts = 2L,
    adjustmentMethod = "holm",
    condNameList = list("origin")
  )

ComplexHeatmap::draw(splitHeatmap)
```

In the above graph, it is possible to see that the found clusters align
well to the expression of the layers' genes.

```{r eval=FALSE}
splitCoexDF <- getClusterizationData(obj, clName = "split")[["coex"]]

enrichment <- geneSetEnrichment(clustersCoex = splitCoexDF,
                                groupMarkers = layersGenes)

enrichmentVals <-
  t(as.matrix(enrichment[, -((ncol(enrichment)-1):ncol(enrichment))]))
stats::heatmap(enrichmentVals, Colv = NA)

tmp <- logFoldChangeOnClusters(obj, clName = "merge", floorLambdaFraction = 0.05)
```

```{r}
tmp <- reorderClusterization(obj, clName = "split")

treePlot <-
  clustersTreePlot(obj, kCuts = 4, clName = "split",
                   useDEA = TRUE, )[["dend"]]

plot(treePlot)
```

### Merged UT clusterization

Loading pre-calculated *clusterization* from data

```{r load-pre-calculated-clusterization_merge}
data("vignette.merge.clusters", package = "COTAN")
mergedClusters <- vignette.merge.clusters[getCells(obj)]

# explicitly calculate the DEA of the clusterization to store it
mergedCoexDF <- DEAOnClusters(obj, clusters = mergedClusters)

obj <-
  addClusterization(
    obj,
    clName = "merge",
    clusters = mergedClusters,
    coexDF = mergedCoexDF,
    override = FALSE
  )

# this gives immediate visual information about which clusters were merged
table(splitClusters, mergedClusters)
```

```{r extract-clusterization-summary_merge}
c(summaryData, summaryPlot) %<-%
  clustersSummaryPlot(
    obj,
    clName = "split",
    plotTitle = "split summary",
    condName = "origin"
  )

head(summaryData, n = 10L)

plot(summaryPlot)
```

### Relevance of marker genes for the `merge` clusters

Again, it is possible to visualize how relevant are the *marker genes* from
above:

```{r  clusterization-data-visualization_merge}
c(mergeHeatmap, mergeScoresDF, mergePValuesDF) %<-%
  clustersMarkersHeatmapPlot(
    obj,
    groupMarkers = layersGenes,
    clName = "merge",
    condNameList = list("origin")
  )

ComplexHeatmap::draw(mergeHeatmap)
```


## Vignette clean-up stage

The next few lines are just to clean.

```{r clean-up}
if (file.exists(file.path(outDir, paste0(cond, ".cotan.RDS")))) {
  # delete file if it exists
  file.remove(file.path(outDir, paste0(cond, ".cotan.RDS")))
}
if (file.exists(file.path(outDir, paste0(cond, "_times.csv")))) {
  # delete file if it exists
  file.remove(file.path(outDir, paste0(cond, "_times.csv")))
}
if (dir.exists(file.path(outDir, cond))) {
  unlink(file.path(outDir, cond), recursive = TRUE)
}
# if (dir.exists(file.path(outDir, GEO))) {
#   unlink(file.path(outDir, GEO), recursive = TRUE)
# }

# stop logging to file
setLoggingFile("")
file.remove(file.path(outDir, "vignette_uniform_clustering.log"))

options(prevOptState)
```

```{r closure}
Sys.time()

sessionInfo()
```
