---
title: "Gene_clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene_clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

library(factoextra)
library(COTAN)
library(ggrepel)
library(Rtsne)
library(plotly)
library(tidyverse)
library(htmlwidgets)
library(MASS)
library(dendextend)
```

To demostrate how to do the gene clustering usign COTAN we begin importing the COTAN object that stores all elaborated data and, in this case, regarding a mouse embrionic cortex dataset (developmental stage E17.5).

```{r}
input_dir = "../../Cotan_paper/Data/"
layers = list("L1"=c("Reln","Lhx5"), "L2/3"=c("Satb2","Cux1"), "L4"=c("Rorb","Sox5") , "L5/6"=c("Bcl11b","Fezf2") , "Prog"=c("Vim","Hes1"))
#objE17 = readRDS(file = paste(input_dir,"E17.5_cortex.cotan.RDS", sep = ""))
objE17 = readRDS(file = paste(input_dir,"E17.5_cortex_cleaned.cotan.RDS", sep = ""))
```

```{r}
g.space = get.gene.coexpression.space(objE17, 
                                      n.genes.for.marker = 28,#25  
                                      primary.markers = unlist(layers))
```
```{r}
g.space = as.data.frame(as.matrix(g.space))

coex.pca.genes <- prcomp(t(g.space),
                 center = TRUE,
                 scale. = F) 

fviz_eig(coex.pca.genes, addlabels=TRUE,ncp = 10)
#fviz_eig(coex.pca.genes, choice = "eigenvalue", addlabels=TRUE)
```
Hierarchical clustering
```{r}

# clustering usign Ward method
hc.norm = hclust(dist(g.space), method = "ward.D2")

# and cut the tree into 5 clusters (for example)
cut = cutree(hc.norm, k = 7, order_clusters_as_data = F)

# It crates the tree
dend <- as.dendrogram(hc.norm)

# I can use a dataframe from the pca to store some data regarding the clustering
pca_1 = as.data.frame(coex.pca.genes$rotation[,1:5])
pca_1 = pca_1[order.dendrogram(dend),]


mycolours <- c("genes related to L5/6" = "#3C5488FF","genes related to L2/3"="#F39B7FFF","genes related to Prog"="#4DBBD5FF","genes related to L1"="#E64B35FF","genes related to L4" = "#91D1C2FF", "not marked"="#B09C85FF")

# save the cluster number into the dataframe
pca_1$hclust = cut

dend =branches_color(dend,k=7,col=c("#4DBBD5FF","#91D1C2FF","#F39B7FFF","#E64B35FF","#3C5488FF","#91D1C2FF","#B09C85FF" ),groupLabels = T)
#dend =color_labels(dend,k=5,labels = rownames(pca_1),col=pca_1$colors)



# plot the dendrogram with all anmes for secondary markers
dend %>%
  set("labels", ifelse(labels(dend) %in% rownames(pca_1)[rownames(pca_1) %in% colnames(as.matrix(g.space))], labels(dend), "")) %>%
  #  set("branches_k_color", value = c("gray80","#4DBBD5FF","#91D1C2FF" ,"gray80","#F39B7FFF","#E64B35FF","#3C5488FF"), k = 7) %>%
 plot(horiz=F, axes=T)#,ylim = c(0,80))

```
or just with primary markers

```{r fig.width= 10}

dend %>%
set("labels", ifelse(labels(dend) %in% rownames(pca_1)[rownames(pca_1) %in% unlist(layers)], labels(dend), "")) %>%
  set("branches_k_color", value = c("gray80","#4DBBD5FF","#91D1C2FF" ,"gray80","#F39B7FFF","#E64B35FF","#3C5488FF"), k = 7) %>%
plot(horiz=F, axes=T,ylim = c(0,100))

```

Now we can plot the PCA
```{r}
# some more genes as landmarks
controls =list("genes related to L5/6"=c("Foxp2","Tbr1"), "genes related to L2/3"=c("Mef2c"), "genes related to Prog"=c("Nes","Sox2") , "genes related to L1"=c() , "genes related to L4"=c()) 

# dataframe to be able to label only primary markers and control genes
textdf <- pca_1[rownames(pca_1) %in% c(unlist(layers),unlist(controls)) , ]

 for (m in c(1:length(controls))) {
  for (g in controls[[m]]) {
    if(g %in% rownames(textdf)){
      textdf[g,"highlight"] = names(controls[m])
    } 
  }
}

# deciding the colors
mycolours <- c("genes related to L5/6" = "#3C5488FF","genes related to L2/3"="#F39B7FFF","genes related to Prog"="#4DBBD5FF","genes related to L1"="#E64B35FF","genes related to L4" = "#91D1C2FF", "not marked"="#B09C85FF")

# to assing correcly the cluster number and the color
mycolours2 = c("Reln","Satb2","Rorb","Bcl11b","Vim")
names(mycolours2) = unique(cut[unlist(layers)])

mycolours2[mycolours2 == "Reln"] = "#E64B35FF"
mycolours2[mycolours2 == "Satb2"] = "#F39B7FFF"
mycolours2[mycolours2 == "Rorb"] = "#91D1C2FF"
mycolours2[mycolours2 == "Bcl11b"] = "#3C5488FF"
mycolours2[mycolours2 == "Vim"] = "#4DBBD5FF"
color_to_add = unique(pca_1$hclust)[!unique(pca_1$hclust) %in% as.numeric(names(mycolours2))]
names(color_to_add) = unique(pca_1$hclust)[!unique(pca_1$hclust) %in% as.numeric(names(mycolours2))]
color_to_add[color_to_add %in% 
                 unique(pca_1$hclust)[!unique(pca_1$hclust) %in% as.numeric(names(mycolours2))]] = "#B09C85FF"
mycolours2 = c(mycolours2,color_to_add)

pca1 = ggplot(subset(pca_1,!hclust %in% unique(cut[unlist(layers)])  ), aes(x=PC1, y=PC2)) +  geom_point(alpha = 0.3,color = "#B09C85FF",size=1)

#pca2 = pca1 + geom_point(data = subset(pca_1, highlight != "not marked" ), aes(x=PC1, y=PC2, colour=hclust),size=2.5,alpha = 0.9) 
pca2 = pca1 + geom_point(data = subset(pca_1, hclust %in% unique(cut[unlist(layers)]) ), aes(x=PC1, y=PC2, colour=as.character(hclust)),size=1,alpha = 0.5) + 
 scale_color_manual( "Status", values = mycolours2)  +
  scale_fill_manual( "Status", values = mycolours2)  +
  xlab("") + ylab("") +
  geom_label_repel(data =textdf , aes(x = PC1, y = PC2, label = rownames(textdf),fill=as.character(hclust)),
                   label.size = NA, 
                   alpha = 0.5, 
                   direction = "both",
                   na.rm=TRUE,
                   seed = 1234) +
  geom_label_repel(data =textdf , aes(x = PC1, y = PC2, label = rownames(textdf)),
                   label.size = NA, 
                   segment.color = 'black',
                   segment.size = 0.5,
                   direction = "both",
                   alpha = 1, 
                   na.rm=TRUE,
                   fill = NA,
                   seed = 1234) +
  ggtitle("PCA") +
  theme_light(base_size=10) +
    theme(axis.text.x=element_blank(),plot.title = element_text(size=14, 
                                    face="italic", 
                                    color="#3C5488FF",
                                    hjust=0.01,
                                    lineheight=1.2,margin = margin(t = 5, b = -15)),
        axis.text.y=element_blank(),
        legend.position = "none")  # titl)

pca2 #+ geom_encircle(data = pca_1, aes(group=hclust)) 
```

t-SNE code and plot

```{r}
# run the t-SNE
cl.genes.tsne = Rtsne(g.space ,initial_dims = 100, dims = 2, perplexity=30,eta = 200, verbose=F, max_iter = 3000,theta=0.4,num_threads = 10,pca_center = T, pca_scale = FALSE, normalize = T )

d_tsne_1 = as.data.frame(cl.genes.tsne$Y)
rownames(d_tsne_1) = rownames(g.space)

d_tsne_1 = d_tsne_1[order.dendrogram(dend),]

# save the cluster numebr inside a dataframe with the t-SNE information
d_tsne_1$hclust = cut

d_tsne_1$names = rownames(d_tsne_1)

# as before to label only some genes
textdf <- d_tsne_1[rownames(d_tsne_1) %in% c(unlist(layers),unlist(controls)),]

for (m in c(1:length(controls))) {
  for (g in controls[[m]]) {
    if(g %in% rownames(textdf)){
      textdf[g,"highlight"] = names(controls[m])
    } 
  }
}


 p1 = ggplot(subset(d_tsne_1,!hclust %in% unique(cut[unlist(layers)])), aes(x=V1, y=V2)) +  geom_point(alpha = 0.3, color = "#B09C85FF", size=1)

p2 = p1 + geom_point(data = subset(d_tsne_1, hclust %in% unique(cut[unlist(layers)]) ), aes(x=V1, y=V2, colour=as.character(hclust)),size=1,alpha = 0.5) +
    scale_color_manual("Status", values = mycolours2)  +
  scale_fill_manual("Status", values = mycolours2)  +
  xlab("") + ylab("")+
  geom_label_repel(data =textdf , aes(x = V1, y = V2, label = names,fill=as.character(hclust)),
                   label.size = NA, 
                   alpha = 0.5, 
                   direction = "both",
                   na.rm=TRUE,
                   seed = 1234) +
  geom_label_repel(data =textdf , aes(x = V1, y = V2, label = names),
                   label.size = NA, 
                   segment.color = 'black',
                   segment.size = 0.5,
                   direction = "both",
                   alpha = 1, 
                   na.rm=TRUE,
                   fill = NA,
                   seed = 1234) +
  ggtitle("t-SNE") +
  theme_light(base_size=10) +
  theme(axis.text.x=element_blank(),plot.title = element_text(size=14, 
                                    face="italic", 
                                    color="#3C5488FF",
                                    hjust=0.01,
                                    lineheight=1.2,margin = margin(t = 5, b = -15)),
        axis.text.y=element_blank(),
        legend.position = "none")  # titl)
p2
```
Code to create an iteractive plot. This can be modified to be used with all the plots.

```{r echo=TRUE}

p = ggplot(d_tsne_1, aes(x=V1, y=V2, text= paste("gene: ",names))) +  
  geom_point(size=2, aes(colour=as.character(hclust)), alpha=0.8) +
  scale_color_manual("Status", values = mycolours2)  +
  xlab("") + ylab("") +
  ggtitle("t-SNE") +
  theme_light(base_size=10) +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

ggplotly(p)
```

Multidimensional scaling (MDS) and plot
```{r}
# run the MDS
genes.dist.euc =  dist(g.space, method =  "euclidean")
#fit <- isoMDS(genes.dist.euc) # not linear
fit <- isoMDS(genes.dist.euc)

fit.genes = as.data.frame(fit$points)

fit.genes = fit.genes[order.dendrogram(dend),]

fit.genes$hclust = cut


fit.genes$names = rownames(fit.genes)

mycolours3 <- c("cluster L5/6 markers" = "#3C5488FF","cluster L2/3 markers"="#F39B7FFF","cluster Prog markers"="#4DBBD5FF","cluster L1 markers"="#E64B35FF","cluster L4 markers" = "#91D1C2FF", "not identified cluster"="#B09C85FF")

#fit.genes$hclust = factor(cutree(hc.norm, 7))
used = vector()
for (k in c(1:length(layers))) {
  #print(k)
  tt =as.numeric(cut[layers[[k]]][1])
  fit.genes[fit.genes$hclust == tt,"cluster"] = paste("cluster",names(layers[k]),"markers", sep = " " )
  used = c(used,cut[layers[[k]]][1])
}

fit.genes[fit.genes$hclust %in% (unique(fit.genes$hclust)[!unique(fit.genes$hclust) %in% used]),]$cluster = "not identified cluster"

textdf <- fit.genes[rownames(fit.genes) %in% c(unlist(layers),unlist(controls)),]

   f1 = ggplot(subset(fit.genes,!hclust %in% unique(cut[unlist(layers)]) ), aes(x=V1, y=V2)) +  geom_point(alpha = 0.3, color = "#B09C85FF", size=1)

f2 = f1 + geom_point(data = subset(fit.genes, hclust %in% unique(cut[unlist(layers)]) ), 
                     aes(x=V1, y=V2, colour=cluster), size=1,alpha = 0.5) +
  scale_color_manual("Status", values = mycolours3)  +
  scale_fill_manual("Status", values = mycolours3)  + 
  xlab("") + ylab("")+
  geom_label_repel(data =textdf , aes(x = V1, y = V2, label = rownames(textdf),fill=cluster),
                   label.size = NA, 
                   alpha = 0.5, 
                   direction ="both",
                   na.rm=TRUE,
                   seed = 1234, show.legend = FALSE) +
  geom_label_repel(data =textdf , aes(x = V1, y = V2, label = rownames(textdf)),
                   label.size = NA, 
                   segment.color = 'black',
                   segment.size = 0.5,
                   direction = "both",
                   alpha = 1, 
                   na.rm=TRUE,
                   fill = NA,
                   seed = 1234, show.legend = FALSE) +
  ggtitle("MDS") +
  theme_light(base_size=10) +
  theme(axis.text.x=element_blank(),plot.title = element_text(size=14, 
                                    face="italic", 
                                    color="#3C5488FF",
                                    hjust=0.01,
                                    lineheight=1.2,margin = margin(t = 5, b = -15)),
        axis.text.y=element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(color = "#3C5488FF",face ="italic" ),
        legend.position = "bottom")  # titl)


f2 #+ geom_encircle(data = fit.genes, aes(group=`5_clusters`)) 
```

```{r}
sessionInfo()
```

