---
title: "Guided tutorial to genes' clustering using COTAN"
author: 
- name: "Silvia Giulia Galfrè"
  affiliation: "Department of Computer Science, University of Pisa"
- name: "Marco Fantozzi"
  affiliation: "University of Parma"

package: COTAN
output:
  BiocStyle::html_document:
    keep_md: true
vignette: >
  %\VignetteIndexEntry{Genes' clustering using COTAN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Preamble

```{r knitr options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7L,
  fig.height = 5L,
  dpi = 100
)

if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
} else if (requireNamespace("Cairo", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "cairo_png")
} else {
  knitr::opts_chunk$set(dev = "png")
}
```


```{r Preamble, message=FALSE, warning=FALSE}
library(COTAN)
library(zeallot)

# necessary to solve precedence of overloads
conflicted::conflict_prefer("%<-%", "zeallot")

# enable multi-processing (not on Windows)
prevOptState <- options(parallelly.fork.enable = TRUE)
```

# Introduction

This tutorial shows the available tools in COTAN to help with genes' clustering

### Setup

Define a directory where the data and the output will be stored

```{r Setup and log}
dataDir <- tempdir()
print(dataDir)

outDir <- dataDir

# Log-level 2 was chosen to showcase better how the package works
# In normal usage a level of 0 or 1 is more appropriate
setLoggingLevel(2L)

# This file will contain all the logs produced by the package
# as if at the highest logging level
setLoggingFile(file.path(outDir, "vignette_genes_clustering.log"))
```


## Retrieving the data-set

Download the data-set for `r "mouse cortex E17.5"` from
`r "Yuzwa et al."` (2017)

```{r Download data, eval=TRUE, include=TRUE}
GEO <- "GSM2861514"

dir.create(file.path(dataDir, GEO), showWarnings = FALSE)

datasetFileName <-
  file.path(dataDir, GEO, "GSM2861514_E175_All_Cells_DGE.txt.gz")
corticalCellsFileName <-
  file.path(dataDir, GEO, "GSM2861514_E175_Only_Cortical_Cells_DGE.txt.gz")

if (!file.exists(datasetFileName)) {
  GEOquery::getGEOSuppFiles(
    GEO,
    makeDirectory = TRUE,
    baseDir = dataDir,
    filter_regex = basename(datasetFileName),
    fetch_files = TRUE
  )
}

rawDataset <- read.csv(datasetFileName, sep = "\t", row.names = 1L)

print(dim(rawDataset))
```

### Create the COTAN object

Initialize the `COTAN` object with the row count table and
the metadata from the experiment.

```{r Create the COTAN object}
cond <- "mouse_cortex_E17.5"

obj <- COTAN(raw = rawDataset)
obj <-
  initializeMetaDataset(
    obj,
    GEO = GEO,
    sequencingMethod = "Drop_seq",
    sampleCondition = cond
  )

logThis(paste0("Condition ", getMetadataElement(obj, datasetTags()[["cond"]])),
        logLevel = 1L)
```

Assign to each cell its origin

```{r Add origing to the COTAn object as a condition}
# mark cells origin
data(vignette.cells.origin)
head(vignette.cells.origin, 18)

obj <- 
  addCondition(obj, condName = "origin", conditions = vignette.cells.origin)
```


### Drop low quality cells

```{r Drop low quality cells}
# use previously established results to determine
# which cells were dropped in the cleaning stage
data(vignette.split.clusters)

cellsToDrop <-
  getCells(obj)[!(getCells(obj) %in% names(vignette.split.clusters))]

obj <- dropGenesCells(obj, cells = cellsToDrop)

# Log the remaining number of cells

logThis(paste("n cells", getNumCells(obj)), logLevel = 1L)

table(getCondition(obj, condName = "origin"))

rm(vignette.split.clusters)
```

# Establish genes' clusters

## COTAN analysis

In this part, the `COTAN` model is calibrated:
all the contingency tables are computed and used to get
the statistics necessary to `COEX` evaluation and storing

```{r Calculate and store COEX matrix}
obj <-
  proceedToCoex(
    obj,
    calcCoex = TRUE,
    optimizeForSpeed = TRUE,
    cores = 3L,
    deviceStr = "cuda"
  )
```

### Get data tables nad COEX used

Sometimes we can also be interested in the numbers present directly in the
contingency tables for two specific genes.

```{r Get the contingency tables and COEX for given gene pair}
cat("Contingency Tables:")
print(contingencyTables(obj, g1 = "Satb2", g2 = "Bcl11b"))

cat(paste("Corresponding COEX value:",
          getGenesCoex(obj, zeroDiagonal = FALSE)["Satb2", "Bcl11b"]))
```

Another useful function is `getGenesCoex()`. This can be used to extract
the whole or a partial `COEX` matrix from a `COTAN` object.

```{r Display small chunk of the COEX matrix}
# Extract the whole matrix
coex <- getGenesCoex(obj, zeroDiagonal = FALSE)
coex[1000L:1005L, 1000L:1005L]
```

```{r Get sub-matrix of the COEX}
# For a partial matrix
coex_sm <- getGenesCoex(obj, genes = c("Satb2", "Bcl11b", "Fezf2"),
                        zeroDiagonal = FALSE)
coex_sm[2001:2010, ]
```

## Analysis of the elaborated data: GDI

To calculate and store the *Global Differentiation Index* (`GDI`)
use the following:

```{r Calculate and store GDI}
gdiDF <- calculateGDI(obj)

# Show the most *differentially expressed* genes
perm <- order(gdiDF[["GDI"]], decreasing = TRUE)
head(gdiDF[perm, ], n = 10)

# This will store only the $GDI column
obj <- storeGDI(obj, genesGDI = gdiDF)
```

### Store the calibrated object for future analysis

```{r Save COTAN obj to file, eval=FALSE}
# saving the structure
saveRDS(obj, file = file.path(outDir, paste0(cond, ".cotan.RDS")))
```




## Vignette clean-up stage

The next few lines are just to clean.

```{r Clean-up}
if (file.exists(file.path(outDir, paste0(cond, ".cotan.RDS")))) {
  # delete file if it exists
  file.remove(file.path(outDir, paste0(cond, ".cotan.RDS")))
}
if (file.exists(file.path(outDir, paste0(cond, "_times.csv")))) {
  # delete file if it exists
  file.remove(file.path(outDir, paste0(cond, "_times.csv")))
}
if (dir.exists(file.path(outDir, cond))) {
  unlink(file.path(outDir, cond), recursive = TRUE)
}
# if (dir.exists(file.path(outDir, GEO))) {
#   unlink(file.path(outDir, GEO), recursive = TRUE)
# }

# stop logging to file
setLoggingFile("")
file.remove(file.path(outDir, "vignette_v2.log"))

options(prevOptState)
```

```{r Closing}
Sys.time()

sessionInfo()
```
