---
title: "Guided tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guided_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(COTAN)
library(data.table)
library(Matrix)
library(ggrepel)
library(latex2exp)
```

```{r}
mycolours <- c("A" = "#8491B4B2","B"="#E64B35FF")
my_theme = theme(axis.text.x = element_text(size = 14, angle = 0, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF" ),
                 axis.text.y = element_text( size = 14, angle = 0, hjust = 0, vjust = .5, face = "plain", colour ="#3C5488FF"),
                 axis.title.x = element_text( size = 14, angle = 0, hjust = .5, vjust = 0, face = "plain", colour ="#3C5488FF"),
                 axis.title.y = element_text( size = 14, angle = 90, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF"))
```


Download the dataset for mouse cortex E17.5.
```{r}
data_dir = "../data/"
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM2861nnn/GSM2861514/suppl/GSM2861514_E175_Only_Cortical_Cells_DGE.txt.gz",
              paste(data_dir,"E175_only_cortical_cells.txt.gz", sep = "/"),method = "wget", quiet = F)
```
```{r}
data = as.data.frame(fread(paste(data_dir,"E175_only_cortical_cells.txt.gz", sep = "/"),sep = "\t"))
data = as.data.frame(data)
rownames(data) = data$V1
data = data[,2:ncol(data)]
data[1:10,1:10]
```

Define a directory where the ouput will be stored.
```{r}
out_dir = "../data/"
```

## Analitical pipeline

Inizialise the COTAN object with the row count table and the metadata for the experiment.

```{r}
obj = new("scCOTAN",raw = data)
obj = initRaw(obj,GEO="GSM2861514" ,sc.method="Drop_seq",cond = "mouse cortex E17.5")
```

Now we can start the cleaning. Analysis requires and starts from a matrix of raw UMI counts after removing possible cell doublets or multiplets and low quality or dying cells (with too high mtRNA percentage, easily done with Seurat or other tools).

If we do not want to consider the mithocondrial genes we can remove them before starting the analysis.
```{r}
genes_to_rem = rownames(obj@raw[grep('^mt', rownames(obj@raw)),]) #genes to remove : mithocondrial
obj@raw = obj@raw[!rownames(obj@raw) %in% genes_to_rem,]
cells_to_rem = colnames(obj@raw[which(colSums(obj@raw) == 0)])
obj@raw = obj@raw[,!colnames(obj@raw) %in% cells_to_rem]
```
We want also to define a prefix to identify the sample.

```{r}
t = "E17.5_cortex"

print(paste("Condition ",t,sep = ""))
#--------------------------------------
n_cells = length(colnames(obj@raw))
print(paste("n cells", n_cells, sep = " "))

n_it = 1

#obj@raw = obj@raw[rownames(cells), colnames(cells)]
```
## Data cleaning
First we create the directory to store all information regardin the data cleaning.
```{r}
if(!file.exists(out_dir)){
  dir.create(file.path(out_dir))
}

if(!file.exists(paste(out_dir,"cleaning", sep = ""))){   
  dir.create(file.path(out_dir, "cleaning"))
}
```

```{r}
ttm = clean(obj)

obj = ttm$object

ttm$pca.cell.2
```

Run this when B cells need to be removed.
```{r}
pdf(paste(out_dir,"cleaning/",t,"_",n_it,"_plots_before_cells_exlusion.pdf", sep = ""))
ttm$pca.cell.2
ggplot(ttm$D, aes(x=n,y=means)) + geom_point() +
  geom_text_repel(data=subset(ttm$D, n > (max(ttm$D$n)- 15) ), aes(n,means,label=rownames(ttm$D[ttm$D$n > (max(ttm$D$n)- 15),])),
                  nudge_y      = 0.05,
                  nudge_x      = 0.05,
                  direction    = "x",
                  angle        = 90,
                  vjust        = 0,
                  segment.size = 0.2)+
  ggtitle("B cell group genes mean expression")+my_theme +
  theme(plot.title = element_text(color = "#3C5488FF", size = 20, face = "italic",vjust = - 5,hjust = 0.02 ))
dev.off()

if (length(ttm$cl1) < length(ttm$cl2)) {
  to_rem = ttm$cl1
}else{
  to_rem = ttm$cl2
}
n_it = n_it+1
obj@raw = obj@raw[,!colnames(obj@raw) %in% to_rem]
#obj@raw = obj@raw[rownames(obj@raw) %in% rownames(cells),colnames(obj@raw) %in% colnames(cells)]
gc()

ttm = clean(obj)
#ttm = clean.sqrt(obj, cells)
obj = ttm$object

ttm$pca.cell.2

```

Run this only in the last iteration, instead the previus code, when B cells group has not to be removed
```{r}
pdf(paste(out_dir,"cleaning/",t,"_",n_it,"_plots_before_cells_exlusion.pdf", sep = ""))
ttm$pca.cell.2
ggplot(ttm$D, aes(x=n,y=means)) + geom_point() +
  geom_text_repel(data=subset(ttm$D, n > (max(ttm$D$n)- 15) ), aes(n,means,label=rownames(ttm$D[ttm$D$n > (max(ttm$D$n)- 15),])),
                  nudge_y      = 0.05,
                  nudge_x      = 0.05,
                  direction    = "x",
                  angle        = 90,
                  vjust        = 0,
                  segment.size = 0.2)+
  ggtitle(label = "B cell group genes mean expression", subtitle = " - B group NOT removed -")+my_theme +
  theme(plot.title = element_text(color = "#3C5488FF", size = 20, face = "italic",vjust = - 10,hjust = 0.02 ),
        plot.subtitle = element_text(color = "darkred",vjust = - 15,hjust = 0.01 ))

dev.off()
```
To color the pca based on nu_j (so the cells' efficiency)

```{r}
nu_est = round(obj@nu, digits = 7)

plot.nu <-ggplot(ttm$pca_cells,aes(x=PC1,y=PC2, colour = log(nu_est)))

plot.nu = plot.nu + geom_point(size = 1,alpha= 0.8)+
  scale_color_gradient2(low = "#E64B35B2",mid =  "#4DBBD5B2", high =  "#3C5488B2" ,
                        midpoint = log(mean(nu_est)),name = TeX(" $ln (\\nu) $ "))+
  ggtitle("Cells PCA coloured by cells efficiency") +
  my_theme +  theme(plot.title = element_text(color = "#3C5488FF", size = 20),
                    legend.title=element_text(color = "#3C5488FF", size = 14,face = "italic"),
                    legend.text = element_text(color = "#3C5488FF", size = 11),
                    legend.key.width = unit(2, "mm"),
                    legend.position="right")

pdf(paste(out_dir,"cleaning/",t,"_plots_PCA_efficiency_colored.pdf", sep = ""))
plot.nu
dev.off()

plot.nu
```
The next part is use to remove the cells with efficiency too low.
```{r}
nu_df = data.frame("nu"= sort(obj@nu), "n"=c(1:length(obj@nu)))

ggplot(nu_df, aes(x = n, y=nu)) + 
    geom_point(colour = "#8491B4B2", size=1)+
    my_theme #+ ylim(0,1) + xlim(0,70)

```

We can zoom on the smalest values and, if we detect a clear elbow, we can decide to remove the cells.
```{r}
yset = 0.38#threshold to remove low UDE cells
plot.ude <- ggplot(nu_df, aes(x = n, y=nu)) + 
    geom_point(colour = "#8491B4B2", size=1) + 
    my_theme + ylim(0,1) + xlim(0,400) +
    geom_hline(yintercept=yset, linetype="dashed", color = "darkred") +
    annotate(geom="text", x=200, y=0.25, 
             label=paste("to remove cells with nu < ",yset,sep = " "), 
             color="darkred", size=4.5)


pdf(paste(out_dir,"cleaning/",t,"_plots_efficiency.pdf", sep = ""))
plot.ude
dev.off()

plot.ude
```
We also save the defined treshold in the metadata and re run the estimation
```{r}
obj@meta[(nrow(obj@meta)+1),1:2] = c("Threshold low UDE cells:",yset)

to_rem = rownames(nu_df[which(nu_df$nu < yset),])
```

Repeat the estimation after the cells are removed

```{r}
ttm = clean(obj)
obj = ttm$object
ttm$pca.cell.2
```

In case we do not want to remove anything, we can run:

```{r}
pdf(paste(out_dir,"cleaning/",t,"_plots_efficiency.pdf", sep = ""))
ggplot(nu_df, aes(x = n, y=nu)) + geom_point(colour = "#8491B4B2", size=1) +my_theme + #xlim(0,100)+
  annotate(geom="text", x=50, y=0.25, label="nothing to remove ", color="darkred")
dev.off()
```

Just to check again, we plot the final efficiency colored PCA

```{r}
nu_est = round(obj@nu, digits = 7)
plot.nu <-ggplot(ttm$pca_cells,aes(x=PC1,y=PC2, colour = log(nu_est)))
plot.nu = plot.nu + geom_point(size = 2,alpha= 0.8)+
  scale_color_gradient2(low = "#E64B35B2",mid =  "#4DBBD5B2", high =  "#3C5488B2" ,
                        midpoint = log(mean(nu_est)),name = TeX(" $ln (\\nu) $ "))+
  ggtitle("Cells PCA coloured by cells efficiency: last") +
  my_theme +  theme(plot.title = element_text(color = "#3C5488FF", size = 20),
                    legend.title=element_text(color = "#3C5488FF", size = 14,face = "italic"),
                    legend.text = element_text(color = "#3C5488FF", size = 11),
                    legend.key.width = unit(2, "mm"),
                    legend.position="right")

pdf(paste(out_dir,"cleaning/",t,"_plots_PCA_efficiency_colored_FINAL.pdf", sep = ""))
plot.nu
dev.off()

plot.nu
```
COTAN analysis: in this part all the contingency tables are computed and used to get the statistics (S)
To storage efficincy of all the observed tables only the yes_yes is stored. 
Note that if will be necessary re-coputing the yes-yes table, the yes-yes table should be cancelled before running cotan_analysis.

```{r}
# obj@yes_yes = c()
obj = cotan_analysis(obj)
```

COEX evaluation and storing 
```{r}
obj = get.coex(obj)

# saving the structure 
saveRDS(obj,file = paste(out_dir,t,".cotan.RDS", sep = ""))
```

## Analysis of the elaborated data

### GDI

The next function can directly plot the GDI for the dataset with the 1.5 threshold (in red) and the two higher quantiles (in blue). 

```{r}
plot.GDI(obj, cond = "E17.5")
```

If a more complex plot is needed, or if we want to analize mor in detail the GDI data, we can get direcly the GDI dataframe. 
```{r echo=TRUE}
quant.p = get.GDI(obj)

head(quant.p)
```
In the third column of this dataframe is reported the percentage of cells expressing the gene.

```{r echo=TRUE}
NPGs=c("Nes","Vim","Sox2","Sox1","Notch1", "Hes1","Hes5","Pax6") #,"Hes3"
PNGs=c("Map2","Tubb3","Neurod1","Nefm","Nefl","Dcx","Tbr1")
hk=c("Calm1","Cox6b1","Ppia","Rpl18","Cox7c","Erh","H3f3a","Taf1b","Taf2","Gapdh","Actb","Golph3", "Mtmr12", "Zfr", "Sub1", "Tars", "Amacr")

text.size = 12

quant.p$highlight = with(quant.p, ifelse(rownames(quant.p) %in% NPGs, "NPGs",
                                                 ifelse(rownames(quant.p) %in% hk,"Constitutive" ,
                                                               ifelse(rownames(quant.p) %in% PNGs,"PNGs" , "normal"))))

textdf <- quant.p[rownames(quant.p) %in% c(NPGs,hk,PNGs), ]
mycolours <- c("Constitutive" = "#00A087FF","NPGs"="#E64B35FF","PNGs"="#F39B7FFF","normal" = "#8491B4B2")
f1 = ggplot(subset(quant.p,highlight == "normal" ), aes(x=sum.raw.norm, y=GDI)) +  geom_point(alpha = 0.1, color = "#8491B4B2", size=2.5)
GDI_plot = f1 +  geom_point(data = subset(quant.p,highlight != "normal"  ), aes(x=sum.raw.norm, y=GDI, colour=highlight),size=2.5,alpha = 0.8) +
  geom_hline(yintercept=quantile(quant.p$GDI)[4], linetype="dashed", color = "darkblue") +
  geom_hline(yintercept=quantile(quant.p$GDI)[3], linetype="dashed", color = "darkblue") +
  geom_hline(yintercept=1.5, linetype="dotted", color = "red", size= 0.5) +
  scale_color_manual("Status", values = mycolours)  +
  scale_fill_manual("Status", values = mycolours)  +
  xlab("log normalized counts")+ylab("GDI")+
  geom_label_repel(data =textdf , aes(x=sum.raw.norm, y=GDI, label = rownames(textdf),fill=highlight),
                   label.size = NA, 
                   alpha = 0.5, 
                   direction ="both",
                   na.rm=TRUE,
                   seed = 1234) +
  geom_label_repel(data =textdf , aes(x=sum.raw.norm, y=GDI, label = rownames(textdf)),
                   label.size = NA, 
                   segment.color = 'black',
                   segment.size = 0.5,
                   direction = "both",
                   alpha = 0.8, 
                   na.rm=TRUE,
                   fill = NA,
                   seed = 1234) +
  theme(axis.text.x = element_text(size = text.size, angle = 0, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF" ),
        axis.text.y = element_text( size = text.size, angle = 0, hjust = 0, vjust = .5, face = "plain", colour ="#3C5488FF"),  
        axis.title.x = element_text( size = text.size, angle = 0, hjust = .5, vjust = 0, face = "plain", colour ="#3C5488FF"),
        axis.title.y = element_text( size = text.size, angle = 90, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF"),
        legend.title = element_blank(),
        legend.text = element_text(color = "#3C5488FF",face ="italic" ),
        legend.position = "bottom")  # titl)
legend <- cowplot::get_legend(GDI_plot)
GDI_plot =GDI_plot + theme(
        legend.position = "none") 
GDI_plot

```

```{r}
list.genes = list("Ref.col"= PNGs, "NPGs"=NPGs, "Const."=hk )

plot.heatmap(df_genes = list.genes,sets = c(1:3),conditions = "E17.5_cortex",dir = "../data/")
```

```{r}
plot.general.heatmap(prim.markers = PNGs, condition = "E17.5_cortex",dir = "../data/", p_value = 0.0001)
```




```{r}
detach("package:data.table", unload = TRUE)
detach("package:Matrix",unload = TRUE)
detach("package:ggrepel", unload = TRUE)
detach("package:latex2exp",unload = TRUE)
```

