---
title: "Guided tutorial to cells Uniform Transcipt cells clustering using COTAN"
author: 
- name: "Silvia Giulia Galfrè"
  affiliation: "Department of Computer Science, University of Pisa"
- name: "Marco Fantozzi"
  affiliation: "University of Parma"

package: COTAN
output:
  BiocStyle::html_document:
    keep_md: true
vignette: >
  %\VignetteIndexEntry{Uniform clustering using COTAN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Preamble

```{r knitr options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7L,
  fig.height = 5L,
  dpi = 100
)

if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
} else if (requireNamespace("Cairo", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "cairo_png")
} else {
  knitr::opts_chunk$set(dev = "png")
}
```


```{r Preamble, message=FALSE, warning=FALSE}
library(COTAN)
library(zeallot)

# necessary to solve precedence of overloads
conflicted::conflict_prefer("%<-%", "zeallot")

# enable multi-processing (not on Windows)
prevOptState <- options(parallelly.fork.enable = TRUE)
```

# Introduction

This tutorial shows the available tools in COTAN to help with genes' clustering

### Setup

Define a directory where the data and the output will be stored

```{r Setup and log}
dataDir <- tempdir()
print(dataDir)

outDir <- dataDir

# Log-level 2 was chosen to showcase better how the package works
# In normal usage a level of 0 or 1 is more appropriate
setLoggingLevel(2L)

# This file will contain all the logs produced by the package
# as if at the highest logging level
setLoggingFile(file.path(outDir, "vignette_uniform_clustering.log"))
```


## Retrieving the data-set

Download the data-set for `r "mouse cortex E17.5"` from
`r "Yuzwa et al."` (2017)

```{r Download data, eval=TRUE, include=TRUE}
GEO <- "GSM2861514"

dir.create(file.path(dataDir, GEO), showWarnings = FALSE)

datasetFileName <-
  file.path(dataDir, GEO, "GSM2861514_E175_All_Cells_DGE.txt.gz")

if (!file.exists(datasetFileName)) {
  GEOquery::getGEOSuppFiles(
    GEO,
    makeDirectory = TRUE,
    baseDir = dataDir,
    filter_regex = basename(datasetFileName),
    fetch_files = TRUE
  )
}

rawDataset <- read.csv(datasetFileName, sep = "\t", row.names = 1L)

print(dim(rawDataset))
```

### Create the COTAN object

Initialize the `COTAN` object with the row count table and
the metadata from the experiment.

```{r Create the COTAN object}
cond <- "mouse_cortex_E17.5"

obj <- COTAN(raw = rawDataset)
obj <-
  initializeMetaDataset(
    obj,
    GEO = GEO,
    sequencingMethod = "Drop_seq",
    sampleCondition = cond
  )

logThis(paste0("Condition ", getMetadataElement(obj, datasetTags()[["cond"]])),
        logLevel = 1L)
```

Assign to each cell its origin

```{r Add origing to the COTAn object as a condition}
# mark cells origin
data(vignette.cells.origin)
head(vignette.cells.origin, 18)

obj <- 
  addCondition(obj, condName = "origin", conditions = vignette.cells.origin)
```


### Drop low quality cells

```{r Drop low quality cells}
# use previously established results to determine
# which cells were dropped in the cleaning stage
data(vignette.split.clusters)

cellsToDrop <-
  getCells(obj)[!(getCells(obj) %in% names(vignette.split.clusters))]

obj <- dropGenesCells(obj, cells = cellsToDrop)

# Log the remaining number of cells

logThis(paste("n cells", getNumCells(obj)), logLevel = 1L)

table(getCondition(obj, condName = "origin"))

rm(vignette.split.clusters)
```

# Establish cells' uniform clusters

## COTAN analysis

In this part, the `COTAN` model is calibrated:
all the contingency tables are computed and used to get
the statistics necessary to `COEX` evaluation and storing

```{r Calculate and store COEX matrix}
obj <-
  proceedToCoex(
    obj,
    calcCoex = TRUE,
    optimizeForSpeed = TRUE,
    cores = 3L,
    deviceStr = "cuda"
  )
```

# Uniform Clustering

It is possible to obtain a cells' *clusterization* based on the concept of 
*uniformity* of expression of the genes across the cells in each cluster.
That is the clusters satisfy the null hypothesis of the `COTAN` model:
the genes expression is not dependent on the cell in consideration.

There are two functions involved into obtaining a proper *clusterization*:
the first is `cellsUniformClustering` that uses standard tools *clusterization*
methods, but then discards and re-clusters any *non-uniform* cluster.

Please note that the most important parameters for the users are the
`GDIThreshold`s inside the **Uniform Transcript** checkers: they define how 
strict is the check. Default constructed advance check gives a pretty strong
guarantee of uniformity for the *cluster*.

```{r Create Uniform-Transcript clusterization, eval=FALSE, include=TRUE}
# This code is a little too computationally heavy to be used in an example
# So we stored the result and we can load it in the next section

# default constructed checker is OK
advChecker <- new("AdvancedGDIUniformityCheck")

c(splitClusters, splitCoexDF) %<-%
  cellsUniformClustering(
    obj,
    initialResolution = 0.8,
    checker = advChecker,
    optimizeForSpeed = TRUE,
    deviceStr = "cuda",
    cores = 6L,
    genesSel = "HGDI",
    saveObj = TRUE,
    outDir = outDir
  )

obj <-
  addClusterization(
    obj,
    clName = "split",
    clusters = splitClusters,
    coexDF = splitCoexDF
  )

table(splitClusters)
```

In the case one already has an existing *clusterization*, it is possible to
calculate the *DEA* `data.frame` and add it to the `COTAN` object.
```{r Load pre-calculated clusterization, eval=TRUE, include=TRUE}
data("vignette.split.clusters", package = "COTAN")
splitClusters <- vignette.split.clusters[getCells(obj)]
splitClusters[is.na(splitClusters)] <- "-1"
names(splitClusters) <- getCells(obj)

splitCoexDF <- DEAOnClusters(obj, clusters = splitClusters)

obj <-
  addClusterization(
    obj,
    clName = "split",
    clusters = splitClusters,
    coexDF = splitCoexDF,
    override = FALSE
  )
```

It is possible to have some statistics about the *clusterization*
```{r Extract clusterization summary}
c(summaryData, summaryPlot) %<-%
  clustersSummaryPlot(
    obj,
    clName = "split",
    plotTitle = "split summary"
  )

summaryData
```

The `ExpGenes` column contains the number of genes that are expressed in any
cell of the relevant *cluster*, while the `ExpGenes25` column contains the number of genes expressed in at the least 25% of the cells of the relevant *cluster*
```{r Summary as plot}
plot(summaryPlot)
```


It is possible to visualize how relevant are the *marker genes'* `lists` with
respect to the given *clusterization*
```{r More complete clusterization data visualization}
c(splitHeatmap, scoreDF, pValueDF) %<-%
  clustersMarkersHeatmapPlot(
    obj,
    groupMarkers = layersGenes,
    clName = "split",
    kCuts = 5L,
    adjustmentMethod = "holm"
  )

ComplexHeatmap::draw(splitHeatmap)
```


Since the algorithm that creates the *clusters* is not directly geared to
achieve cluster uniformity, there might be some *clusters* that can be merged
back together and still be **uniform**.

This is the purpose of the function `mergeUniformCellsClusters` that takes a
*clusterization* and tries to merge cluster pairs after checking that together
the pair forms a uniform cluster.

In order to avoid running the totality of the possible checks (as it can explode
quickly with the number of *clusters*), the function relies on a *related*
distance the find the cluster pairs that have the highest chance to be merged.

```{r Merge potentially Uniform-Transcript clusters, eval=FALSE, include=TRUE}
# default constructed checker is OK
advChecker <- new("AdvancedGDIUniformityCheck")

c(mergedClusters, mergedCoexDF) %<-%
  mergeUniformCellsClusters(
    obj,
    clusters = splitClusters,
    checkers = advChecker,
    optimizeForSpeed = TRUE,
    deviceStr = "cuda",
    cores = 6L,
    saveObj = TRUE,
    outDir = outDir
  )

# merges are:
#  1 <- 06 + 07
#  2 <- '-1' + 08 + 11
#  3 <- 09
#  4 <- 01
#  5 <- 02
#  6 <- 12
#  7 <- 03 + 04 + 10 + 13
#  8 <- 05

obj <-
  addClusterization(
    obj,
    clName = "merge",
    override = TRUE,
    clusters = mergedClusters,
    coexDF = mergedCoexDF
  )

table(mergedClusters)
```

Again, here, the most important parameters for the users are the `GDIThreshold`s
inside the **Uniform Transcript** checkers: they define how  strict is the
check. Default constructed advance check gives a pretty strong guarantee of
uniformity for the *cluster*. If one wants to achieve a more *relaxed* merge,
it is possible to define a sequence of checkers, each less stringent than the
previous, that will be applied sequentially: First all the merges with the
stricter checker are performed, than those that satisfy the second, and so on.

```{r Load pre-calculated merged clusterization, eval=FALSE, include=TRUE}

checkersList <-
  list(
    advChecker,
    shiftCheckerThresholds(advChecker, 0.01),
    shiftCheckerThresholds(advChecker, 0.03)
  )

prevCheckRes <- data.frame()

# In this case we want to re-use the already calculated merge checks
# Se we reload them from the output files. This, along a similar facility for
# the split method, is also useful in the cases the execution is interrupted
# prematurely...
#
if (TRUE) {
  # read from the last file among those named all_check_results_XX.csv
  mergeDir <- file.path(outDir, cond, "leafs_merge")
  resFiles <-
    list.files(
      path = mergeDir,
      pattern = "all_check_results_.*csv",
      full.names = TRUE
    )

  prevCheckRes <-
    read.csv(resFiles[length(resFiles)], header = TRUE, row.names = 1L)
}

c(mergedClusters2, mergedCoexDF2) %<-%
  mergeUniformCellsClusters(
    obj,
    clusters = splitClusters,
    checkers = checkersList,
    allCheckResults = prevCheckRes,
    optimizeForSpeed = TRUE,
    deviceStr = "cuda",
    cores = 6L,
    saveObj = TRUE,
    outDir = outDir
  )

# merges are:
#  1 <- '-1' + 06 + 09
#  2 <- 07
#  3 <- 03  + 04 + 10 + 13
#  4 <- 12
#  5 <- 01 + 08 + 11
#  6 <- 02 + 05

obj <-
  addClusterization(
    obj,
    clName = "merge2",
    override = TRUE,
    clusters = mergedClusters2,
    coexDF = mergedCoexDF2
  )

table(mergedClusters2)
```


In the case one already has existing *clusterizations*, it is possible to
calculate their *DEA* `data.frame` and add them to the `COTAN` object.
```{r Calculate and store DEA using clusterizations, eval=FALSE}
data("vignette.merge.clusters", package = "COTAN")
mergedClusters <- vignette.merge.clusters[getCells(obj)]

mergedCoexDF <- DEAOnClusters(obj, clusters = mergedClusters)

obj <-
  addClusterization(
    obj,
    clName = "merge",
    clusters = mergedClusters,
    coexDF = mergedCoexDF,
    override = FALSE
  )

data("vignette.merge2.clusters", package = "COTAN")
mergedClusters2 <- vignette.merge2.clusters[getCells(obj)]

mergedCoexDF2 <- DEAOnClusters(obj, clusters = mergedClusters2)

obj <-
  addClusterization(
    obj,
    clName = "merge2",
    clusters = mergedClusters2,
    coexDF = mergedCoexDF2,
    override = FALSE
  )

table(mergedClusters2, mergedClusters)
```


Here is possible to visualize how the `split` clusters are merged in to 
`merge` and `merge2`

```{r  Clusterizations data visualization, eval=FALSE}
c(mergeHeatmap, mergeScoresDF, mergePValuesDF) %<-%
  clustersMarkersHeatmapPlot(
    obj,
    clName = "merge",
    condNameList = "split",
    conditionsList = list(splitClusters)
  )

ComplexHeatmap::draw(mergeHeatmap)


c(merge2Heatmap, merge2ScoresDF, merge2PValuesDF) %<-%
  clustersMarkersHeatmapPlot(
    obj,
    clName = "merge2",
    condNameList = "split",
    conditionsList = list(splitClusters)
  )

ComplexHeatmap::draw(merge2Heatmap)
```

In the following graph, it is possible to check that the found clusters align
very well to the expression of the layers' genes defined above...

It is possible to plot the *clusterization* on a `UMAP` plot

```{r Clusterization UMAP plot using Dimensionality Reduction, eval=FALSE}
c(umapPlot, cellsRDM) %<-%
  cellsUMAPPlot(
    obj,
    clName = "merge2",
    clusters = NULL,
    useCoexEigen = TRUE,
    dataMethod = "LogLikelihood",
    numComp = 50L,
    genesSel = "HGDI",
    numGenes = 200L,
    colors = NULL,
    numNeighbors = 30L,
    minPointsDist = 0.3
  )

plot(umapPlot)
```


## Vignette clean-up stage

The next few lines are just to clean.

```{r Clean-up}
if (file.exists(file.path(outDir, paste0(cond, ".cotan.RDS")))) {
  # delete file if it exists
  file.remove(file.path(outDir, paste0(cond, ".cotan.RDS")))
}
if (file.exists(file.path(outDir, paste0(cond, "_times.csv")))) {
  # delete file if it exists
  file.remove(file.path(outDir, paste0(cond, "_times.csv")))
}
if (dir.exists(file.path(outDir, cond))) {
  unlink(file.path(outDir, cond), recursive = TRUE)
}
# if (dir.exists(file.path(outDir, GEO))) {
#   unlink(file.path(outDir, GEO), recursive = TRUE)
# }

# stop logging to file
setLoggingFile("")
file.remove(file.path(outDir, "vignette_v2.log"))

options(prevOptState)
```

```{r Closing}
Sys.time()

sessionInfo()
```
